---
title: "final_project_muxin"
author: "Muxin Li"
date: "2025-12-03"
output: html_document
---

Behavior pattern based on age ranges  2019 vs 2020   (code)
Riding distance plot 2019-2020 per month (code)

```{r}
library(tidyverse)
library(geosphere)
```

```{r}
bike_df=read.csv("data/bike_data_combined.csv")

bike_year_distance=
  bike_df|>
  mutate(
    age=year-birth_year,
    distance_m = distHaversine(
      cbind(start_station_longitude, start_station_latitude),
      cbind(end_station_longitude,   end_station_latitude)))|>
  mutate(distance_km = distance_m / 1000)|>
  mutate(
    age_group = cut(
      age,
      breaks = c(-Inf, 17, 29, 49, 64, Inf),
      labels = c("<18", "18–29", "30–49", "50–64", "65+"),
      right = TRUE
    )
  )|>
  filter(age_group!="<18"& age_group!="65+")|>
  select(year,month,tripduration,age_group,distance_km,distance_m)
      
```

```{r}
  
bike_time=bike_year_distance|>
  mutate(year = factor(year),
         month = factor(
      month,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                 "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    ))|>
  group_by(year,month,age_group)|>
  summarise(time=mean(tripduration))

ggplot(bike_time, 
       aes(x = age_group, y = time, color = year, group = year)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.8) +
  facet_wrap(~ month, ncol = 4) +
  labs(
    title = "Monthly Comparison of Riding Time Across Age Groups (2019 vs 2020)",
    x = "Age Group",
    y = "Average Riding Time (seconds)",
    color = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1)
  )
```


```{r}
ggplot(
  bike_time,
  aes(x = month, y = time, color = age_group, group = age_group)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  facet_grid(. ~ year) + 
  labs(
    title = "Monthly Riding Time by Age Group (2019 vs 2020)",
    x = "Month",
    y = "Average Riding Time (seconds)",
    color = "Age Group"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1)
  )
```



12个月对比
```{r}
bike_distance=bike_year_distance|>
  filter(distance_km<100 & distance_km>0)|>
  mutate(year = factor(year),
         month = factor(
      month,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                 "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    ))|>
  group_by(year,month,age_group)|>
  summarise(distance=mean(distance_m))

ggplot(bike_distance,
       aes(x = age_group, y = distance, color = year, group = year)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.8) +
  facet_wrap(~ month, ncol = 4) +
  labs(
    title = "Monthly Comparison of Riding Distance Across Age Groups (2019 vs 2020)",
    x = "Age Group",
    y = "Average Riding Distance (m)",
    color = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1)
  )

```

```{r}
bike_distance=
  bike_distance|>
  filter(age_group!="<18"& age_group!="65+")

ggplot(
  bike_distance,
  aes(x = month, y = distance, color = age_group, group = age_group)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  facet_wrap(~ year, nrow = 1) +  
  labs(
    title = "Monthly Riding Distance by Age Group (2019 vs 2020)",
    x = "Month",
    y = "Average Riding Distance (m)",
    color = "Age Group"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1)
  )

```

summary:
Beginning in March–April 2020, both riding distance and riding time increased sharply across all age groups compared with 2019. This shift aligns with the early COVID-19 outbreak, when PCR testing requirements and the high infection risk on public transportation led many people to avoid subways and buses. As a result, cycling became a safer and more reliable way to commute, travel to testing sites. These pandemic-driven changes persisted throughout the rest of 2020.